Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-03-25 analysis/paper/paper.Rmd
FINAL EDITS BEFORE SHARING W/ CO-AUTH

Ice-associated seals rely on sea ice for a variety of activities, including breeding,
molting, pupping, and resting. In the Arctic, many of these activities occur in spring and
early summer (April-June) as sea ice begins to melt and retreat northward. Rapid
acceleration of climate change in Arctic ecosystems is thus of concern, as the quantity
and quality of suitable habitat is forecast to decrease. In this paper, we use data
collected from satellite-linked telemetry tags deployed between 2005 and 2020 to
investigate the seasonal timing and environmental factors affecting sea ice use by seals
(specifically, bearded, ribbon, and spotted seals) in the Bering, and Chukchi seas. In
addition to providing baseline data on phenology, these data also allow us to refine
availability estimates needed to accurately estimate abundance from aerial survey counts
of seals basking on ice (i.e., to correct for the proportion of animals that are in the
water while surveys are conducted). Using generalized linear mixed pseudo-models to
properly account for temporal autocorrelation, we fit models with a variety of covariates
(e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure,
temperature, precipitation) to examine their ability to explain variation in hourly
haul-out records. We found evidence for strong diurnal and within season patterns in
haul-out behavior, as well as strong weather effects (particularly wind and temperature).
In general, seals were more likely to bask on ice in the middle of the day and when wind
speed was low and temperatures were higher. Haul-out propensity increased through March
and April, peaking in May and early June before declining again. The timing and frequency
of haul-out events also varied substantially based on species and age-sex class. For
ribbon and spotted seals, models with year effects were highly supported, indicating that
the timing and magnitude of haul-out behavior varied among years. Our
analysis also emphasizes the importance of accounting for factors affecting haul-out
behavior when interpreting the number of seals counted in aerial surveys.

@@

```{r setup, include = FALSE}
library(tidyverse)
library(lubridate)
library(glue)
library(DBI)
library(dbplyr)
library(sf)
library(flextable)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(glmmLDTS)
library(mgcv)
library(ragg)

@@

@@

In addition to sex and age class (non-dependent *young of the year*, sexually immature
*subadults*, and mature *adults*), we gathered additional variables that might help
explain variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and solar hour (for diurnal effects). Solar hour was calculated using the {solaR}
package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019] using the
daily locations. We also compiled several meteorological variables shown to affect
haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz]. In
particular, we linked locations from satellite tags to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
data to produce a long-term picture of weather over North America. Numerous weather
variables are made available across the region 8 times daily. For this study, NARR weather
values were subset to the extent of our study area over the Bering and Chukchi Seas at
3-hr intervals based on a grid resolution of 32 km (1024 sq. km) The following
meteorological variables were interpolated and assigned to seal locations using a bilinear
method: 1) air temperature at 2m above the earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the Euclidean norm,
3) barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

@@

We did explore whether the annual variation in spring sea ice extent in the Bering Sea
influences the peak timing of seal haul-out behavior. For this, all sea ice concentration
grid cells (25 $km^2$) in the study area with greater than 15% concentration were summed to
get total sea ice extent for each day between 15 February and 15 July across all years 
(`r min(lubridate::year(dat.sf$haulout_dt))`-`r max(lubridate::year(dat.sf$haulout_dt))`). 
The maximum sea ice extent for each year was determined from this calculation. We obtained
sea ice concentration data from the Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave
Dataset, Version 1 [@cavalieri1996].

@@

@@

Ribbon seals show a pattern of gradually increasing haul-out probability in April that
peaks in May for subadults and in June for adults. The behavior is clearly centered
around local solar noon and expands to other hours later in the season as seals enter
their molting period. subadults showed an earlier start and more intense haul-out
activity in April and May. The young of the year records begin after weaning and the model
predictions demonstrate the ontogeny of in-water activities (e.g. diving, foraging) in
May. Adult females have a more protracted season compared to males and more time spent
resting in June and July. The model suggests adult male ribbon seals complete their molt
by the end of June. Lastly, for adults of both sexes there is some indication for a shift
to a nocturnal haul-out pattern in late June and July.

@@

The haul-out probability for ribbon seals was strongly influenced by each of the weather
covariates included within the model. The impact of temperature, wind, and precipitation
was especially noticeable as ribbon seals were less likely to haul out at higher winds and
higher precipitation rates and more likely to be on the ice when temperatures were
relatively warm. Wind chill (temperature $\times$ wind) also had a negative influence on haul-out
probability.

```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Predicted haul-out probability of ribbon seals across a range of weather conditions."}
library(ggsci)

@@

Spotted seal haul-out behavior appears most strongly influenced by temperature and wind.
Spotted seals were less likely to haul out at higher winds and more likely to be on the
ice when temperatures were relatively warm. Wind chill (temperature $\times$ wind) also had a
negative influence on haul-out probability. Barometric pressure and precipitation were
influential to a slightly lesser extent. Differences in the magnitude of response between
the age-sex classes is present and consistent across each of the weather covariates.

@@

Like ribbon and spotted seals, bearded seal haul-out probability appears influenced by
wind and temperature. And, the model also predicted a response to wind chill. Any
influence of barometric pressure or precipitation is less apparent.

```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Predicted haul-out probability of bearded seals across a range of weather conditions."}
library(ggsci)

@@

<!-- The second set of models that included annual variation in haul-out patterns produced -->
<!-- p-values that were strongly significant for linear (ribbon seals: F10,76662=9.92, -->
<!-- p\<0.0001; spotted seals: F8,51981=5.23, p\<0.0001) and quadratic (ribbon seals: -->
<!-- F10,76662=8.02, p\<0.0001; spotted seals: F8,51981=3.79, p=0.0002) year-by-day -->
<!-- interactions. Predicted distributions of haul-out activity were largely unimodal, but -->
<!-- varied substantially among and within years with respect to both the timing and magnitude -->
<!-- of haul-out peaks (Figure \@ref(fig:annualHO)). For instance, in 2010 and 2016 spotted -->
<!-- seal haul-out activity peaked in early May, while in other years (e.g. 2011) it appeared -->
<!-- to peak three weeks later. It is important to note that predicted variation in annual -->
<!-- haul-out patterns likely reflects both process error and sampling variability. For -->
<!-- instance, some of the more extreme patterns (e.g. ribbon seals in 2012 , spotted seals in -->
<!-- 2014) were for years with only one tagged seal and are unlikely to represent -->
<!-- population-level processes. -->

@@

The timing of peak haul-out probability for ribbon and spotted seals appeared to have 
only a limited to no relationship with the amount of yearly maximum sea ice extent. Adult female and
subadult spotted seals show a negative trend line but no indication that the observed
trend is meaningful (R-squared and p-values for adult female spotted seals 
($R^{2}$: `r formatC(pl_adultF$r.squared, digits=3)`; $p$: 
`r formatC(pl_adultF$p.value, digits=3)`) and
spotted seal subadults($R^{2}$: `r formatC(pl_subadult$r.squared, digits=3)`; 
$p$: `r formatC(pl_subadult$p.value, digits=3)`)). 
For ribbon seals and adult male spotted seals, p-values were
substantially larger than 0.05 (R-squared and p-values for ribbon seals (adult
females($R^{2}$: `r formatC(hf_adultF$r.squared, digits=3)`; $p$: 
`r formatC(hf_adultF$p.value, digits=3)`), 
adult males($R^{2}$: `r formatC(hf_adultM$r.squared, digits=3)`; $p$: 
`r formatC(hf_adultM$p.value, digits=3)`), 
subadults($R^{2}$: `r formatC(hf_subadult$r.squared, digits=3)`; $p$: `r formatC(hf_subadult$p.value, digits=3)`)) and 
spotted seals (adult males($R^{2}$: `r formatC(pl_adultM$r.squared, digits=3)`; $p$: `r formatC(pl_adultM$p.value, digits=3)`)).

@@

@@

In this paper, we fit statistical models to data from sensors on satellite tagged
ice-associated seals in order to examine factors affecting the probability that Arctic
seals haul out on ice. Such probabilities are important for understanding the timing and
ecology of seal behavior, since haul out peaks are associated with important life history
functions such as breeding, pupping and molting. Knowledge of factors affecting haul-out
probabilities is a first step in being able to predict possible impacts of losses of sea
ice on ice-associated seals. Understanding the factors affecting haul-out probabilities is
also necessary to properly interpret seal counts made during aerial surveys. For instance,
a count might be low because there are few seals, or because the timing and conditions are
such that many seals are in the water.

Our analysis revealed that seals use sea ice progressively more throughout the spring, and
with peak use near mid-May to early June before declining again. We also found that seals
preferentially haul out on ice shortly after solar noon, which may be associated with
fewer prey in the water column (??? Can someone else who knows more about their feeding
ecology contribute here?). Interestingly, bearded seals appeared to have two peaks in
haul-out activity throughout the day, one shortly after solar noon, and one centered near
solar midnight (related to benthic feeding?). It appeared that adult spotted seals
initiated haul-out behavior earlier in the year than ribbon seals, and maintained this
behavior over a longer duration; however they exhibited stronger diel behavior than ribbon
seals (Fig X). These patterns are similar to those obtained by Conn et al. (2014), who
used a subset of the data we analyzed here.

Unlike previous analyses of seal haul-out data (e.g. Ver Hoef et al. 2009, Conn et al.
2014), we also investigated the influence of sex-age class on haul-out probabilities.
Although it was difficult to discern a pattern for bearded seals owing to low sample size,
it appeared that adult male spotted and ribbon seals typically hauled out for longer
durations, perhaps because they were less constrained by energetic demands associated with
nursing and could sustain longer periods on ice to engage in molting and mating
activities. Subadult ribbon and spotted seals exhibited different behavior when compared
to adults. In particular, subadult ribbon seals exhibited similar haul-out behavior to
adults, whereas subadult spotted seals appeared to haul-out less frequently and sustained
a shorter duration of peak haul-out behavior than adults. The latter result is likely due
to subadults only needing to haul out on ice to molt; pupping and mating are not yet in
their behavioral repertoire.

We also investigated the influence of environmental covariates on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind chill.
These have been investigated for walrus (e.g. Udevitz et al. 2009), but to our knowledge
have never been investigated for ice-associated seals. In particular, we found clear
evidence that seal haul-out probabilities decrease with wind speed and increase with
temperature (Figs X-Y). In addition, we found evidence that ribbon seal haul-out
probabilities are positively related to barometric pressure, and that spotted seal
haul-out probabilities are negatively related to precipitation (Appendix C). The tendency
of seals to haul out in "good" weather may be related to energetic considerations, better
predator detection, or both. Bearded seal haul-out behavior was predicted to change
slightly based on how far north tagged seals were; seals farther north tended to initiate
haul-out behavior slightly later, and for a longer period of time (Fig. X). This response
may be related to different temperature regimes, or to timing of ice melt, which occurs
later in locations farther north.

There appeared to be rather substantial annual deviations in the timing and magnitude of
haul-out peaks (Figs X-Y). For example, peak ribbon seal haul-out behavior appeared to
vary by as much as three weeks between seasons. There thus seems to be a reasonable amount
of plasticity in the timing of haul-out, suggesting some plasticity in life history
functions such as pupping, molting and breeding. However, we were unable to relate the
timing of haul-out conditions to underlying ice conditions, which makes predicting climate
change consequences on life history events difficult. Future investigation into
environmental factors contributing to the timing of haul-out peaks seems warranted. For
instance, Ray et al. (2010) suggested that ribbon and spotted seals were most dependent on
loose pack ice at the sourthern sea ice edge; thus, the relative suitability for haul-out
behavior may have more to do with the characterization, position, and quantity of loose
pack ice rather than the total amount of ice in the Bering Sea. Further work to
characterize remotely sensed sea ice data into relevant ice types (e.g. loose pack,
continuous pack, rounded pack) would be useful.

Previous attempts to estimate the abundance of phocid seals from aerial survey data in the
Bering and Chukchi Seas (e.g. Bengtson et al. 2005, Conn et al. 2014, Ver Hoef et al.
2014) have used estimated haul-out probabilities to correct for the proportion of animals
that are in the water and thus unavailable to be counted during aerial surveys. Although
several of these studies allowed haul-out probabilities to vary by day-of-year and
time-of-day, previous correction factors have not accounted for variability between years,
different weather conditions, and the age-sex class of the sample. In this paper, we have
shown that there can be considerable differences in the number of seals hauled out on ice
based on these factors. We thus recommend that future abundance analyses employ correction
factors that account for these factors. For instance, it is relatively straightforward to
obtain weather reanalysis products for times and locations that are surveyed and to
construct a relevant correction factor based on predictions of GMPMs. Further, one can
apply stable stage distribution estimates from matrix population models to adjust
predictions to the "average" prediction of the population. The latter suggestion is
particularly appropriate when the age structure of the tagged sample does not reflect the
likely age structure of the population. For instance, population weighted estimates can
differ substantially from those based on a naïve analysis of the tagged sample, as was the
case with analysis of spotted seal data (e.g. Fig X). The most challenging element in
developing availability correction factors is with annual variability. It can be difficult
to get a sufficient sample size to estimate year-specific correction factors, particularly
because research teams would likely need to tag seals and conduct aerial surveys
concurrently which requires considerably more personnel and money. One possible suggestion
is to estimate a "shift" parameter within models for aerial survey counts that allow the
peak of haul-out distributions to be adjusted earlier or later in the year based on the
frequency of counts observed over time. Regardless, researchers should anticipate there
being some unmodeled heterogeneity in availability probability present in abundance
estimates obtained from aerial surveys. This may make trend detection difficult, as one
will not know if moderate differences in abundance estimates are attributable to changes
in abundance or changes in haul-out frequencies.

>>>>

Ice-associated seals rely on sea ice for a variety of activities, including breeding,
molting, pupping, and resting. In the Arctic, many of these activities occur in spring and
early summer (April-June) as sea ice begins to melt and retreat northward. Rapid
acceleration of climate change in Arctic ecosystems is thus of concern as the quantity and
quality of suitable habitat is forecast to decrease. In this paper, we use data collected
from satellite-linked telemetry tags deployed between 2005 and 2020 to investigate the
seasonal timing and environmental factors affecting sea ice use by seals (specifically,
bearded, ribbon, and spotted seals) in the Bering, and Chukchi seas. In addition to
providing baseline data on phenology, these data also allow us to refine availability
estimates needed to accurately estimate abundance from aerial survey counts of seals
basking on ice (i.e., to correct for the proportion of animals that are in the water while
surveys are conducted). Using generalized linear mixed pseudo-models to properly account
for temporal autocorrelation, we fit models with a variety of covariates (e.g.,
day-of-year, solar hour, age-sex class, wind speed, barometric pressure, temperature,
precipitation) to examine their ability to explain variation in hourly haul-out records.
We found evidence for strong diurnal and within season patterns in haul-out behavior, as
well as strong weather effects (particularly wind and temperature). In general, seals were
more likely to rest on ice in the middle of the day and when wind speed was low and
temperatures were higher. Haul-out propensity increased through March and April, peaking
in May and early June before declining again. The timing and frequency of haul-out events
also varied based on species and age-sex class. For ribbon and spotted seals, models with
year effects were highly supported, indicating that the timing and magnitude of haul-out
behavior varied among years. There was no support linking the variability in haul-out
timing to annual sea ice extent. Our analysis also emphasizes the importance of accounting
for weather factors affecting haul-out behavior when interpreting the number of seals
counted in aerial surveys.

@@

```{r setup, include = FALSE}
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(DBI)
library(dbplyr)
library(sf)
library(flextable)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(glmmLDTS)
library(mgcv)
library(ragg)

@@

options(scipen = 3, digits = 3)

@@

In addition to sex and age class (non-dependent *young of the year*, sexually immature
*subadults*, and mature *adults*), we gathered additional variables that might help
explain variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and solar hour (for diurnal effects). Solar hour was calculated using the {solaR}
package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019] using the
daily locations. We also compiled several meteorological variables shown to affect
haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz; @perry2017]. In
particular, we linked locations from satellite tags to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
data to produce a long-term picture of weather over North America. Numerous weather
variables are made available across the region 8 times daily. For this study, NARR weather
values were subset to the extent of our study area over the Bering and Chukchi Seas at
3-hr intervals based on a grid resolution of 32 km (1024 sq. km) The following
meteorological variables were interpolated and assigned to seal locations using a bilinear
method: 1) air temperature at 2m above the earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the Euclidean norm,
3) barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

@@

We did explore whether the annual variation in spring sea ice extent in the Bering Sea
influences the peak timing of seal haul-out behavior. For this, all sea ice concentration
grid cells (25 km^2^) in the study area with greater than 15% concentration were summed to
get total sea ice extent for each day between 15 February and 15 July across all years
(`r min(lubridate::year(dat.sf$haulout_dt))`-`r max(lubridate::year(dat.sf$haulout_dt))`).
The maximum sea ice extent for each year was determined from this calculation. We obtained
sea ice concentration data from the Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave
Dataset, Version 1 [@cavalieri1996].

@@

::: {custom-style="comment"}
I didn't update this at all from Paul's previous text other than to revise some of the
latex equations and add references. Would be good to confirm all of this is still
appropriate and makes sense.
:::

@@

Ribbon seals show a pattern of gradually increasing haul-out probability in April that
peaks in May for subadults and in June for adults. The behavior is clearly centered around
local solar noon and expands to other hours later in the season as seals enter their
molting period. Subadults showed an earlier start and more intense haul-out activity in
April and May. The young of the year records begin after weaning and the model predictions
demonstrate the ontogeny of in-water activities (e.g. diving, foraging) in May. Adult
females have a more protracted season compared to males and more time spent resting in
June and July. The model suggests adult male ribbon seals complete their molt by the end
of June. Lastly, for adults of both sexes there is some indication for a shift to a
nocturnal haul-out pattern in late June and July.

@@

```{r}
ribbon_type3 <- tibble(
  fit_ribbon$typeIII.hypoth
) %>% 
  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>% 
  column_to_rownames(var = "Effect") 
ribbon_type3 <- setNames(split(ribbon_type3, 
                               seq(nrow(ribbon_type3))), 
                         rownames(ribbon_type3))

spotted_type3 <- tibble(
  fit_spotted$typeIII.hypoth
) %>% 
  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
spotted_type3 <- setNames(split(spotted_type3, 
                               seq(nrow(spotted_type3))), 
                         rownames(spotted_type3))
bearded_type3 <- tibble(
  fit_bearded$typeIII.hypoth
) %>% 
  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
bearded_type3 <- setNames(split(bearded_type3, 
                               seq(nrow(bearded_type3))), 
                         rownames(bearded_type3))
```

The haul-out probability for ribbon seals was strongly influenced by each of the weather
covariates included within the model except precipitation. The impact of temperature
($F_{`r paste(ribbon_type3$temp2$Num.df, ribbon_type3$temp2$Den.df, sep=",")`}$ = 
`r ribbon_type3$temp2$F.value`; $p$ = `r ribbon_type3$temp2$Prob.F`), wind
($F_{`r paste(ribbon_type3$wind$Num.df, ribbon_type3$wind$Den.df, sep=",")`}$ = 
`r ribbon_type3$wind$F.value`; $p$ = `r ribbon_type3$wind$Prob.F`), and barometric pressure
($F_{`r paste(ribbon_type3$pressure$Num.df, ribbon_type3$pressure$Den.df, sep=",")`}$ = 
`r ribbon_type3$pressure$F.value`; $p$ = `r ribbon_type3$pressure$Prob.F`) was especially
noticeable as ribbon seals were less likely to haul out at higher winds and lower pressure
values and more likely to be on the ice when temperatures were relatively warm. Wind chill
($F_{`r paste(ribbon_type3$temp2_wind$Num.df, ribbon_type3$temp2_wind$Den.df, sep=",")`}$ 
= `r ribbon_type3$temp2_wind$F.value`; $p$ = `r ribbon_type3$temp2_wind$Prob.F`) showed a
moderately negative influence on haul-out probability.

```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Predicted haul-out probability of ribbon seals across a range of weather conditions."}

@@

Spotted seal haul-out behavior appears most strongly influenced by barometric pressure
($F_{`r paste(spotted_type3$pressure$Num.df, spotted_type3$pressure$Den.df, sep=",")`}$ = 
`r spotted_type3$pressure$F.value`; $p$ = `r spotted_type3$pressure$Prob.F`) and wind
($F_{`r paste(spotted_type3$wind$Num.df, spotted_type3$wind$Den.df, sep=",")`}$ =
`r spotted_type3$wind$F.value`; $p$ = `r spotted_type3$wind$Prob.F`) with temperature
having a moderate influence
($F_{`r paste(spotted_type3$temp2$Num.df, spotted_type3$temp2$Den.df, sep=",")`}$ =
`r spotted_type3$temp2$F.value`; $p$ = `r spotted_type3$temp2$Prob.F`). Spotted seals were
less likely to haul out at higher winds and more likely to be on the ice when temperatures
were relatively warm. Wind chill (temperature $\times$ wind) and precipitation were not as
influential. Differences in the magnitude of response between the age-sex classes is
present and consistent across each of the weather covariates.

@@

Like ribbon and spotted seals, bearded seal haul-out probability appears most influenced
by wind ($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$ =
`r bearded_type3$wind$F.value`; $p$ = `r bearded_type3$wind$Prob.F`) and temperature
($F_{`r paste(bearded_type3$temp2$Num.df, bearded_type3$temp2$Den.df, sep=",")`}$ =
`r bearded_type3$temp2$F.value`; $p$ = `r bearded_type3$temp2$Prob.F`). Additionally, wind
chill was also a significant influence
($F_{`r paste(bearded_type3$temp2_wind$Num.df, bearded_type3$temp2_wind$Den.df, sep=",")`}$ 
= `r bearded_type3$temp2_wind$F.value`; $p$ = `r bearded_type3$temp2_wind$Prob.F`). Any
influence of barometric pressure or precipitation is less apparent.

```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Predicted haul-out probability of bearded seals across a range of weather conditions."}

@@

```{r}
ribbon_year_type3 <- tibble(
  fit_ribbon_year$typeIII.hypoth
) %>% 
  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>% 
  column_to_rownames(var = "Effect") 
ribbon_year_type3 <- setNames(split(ribbon_year_type3, 
                               seq(nrow(ribbon_year_type3))), 
                         rownames(ribbon_year_type3))

spotted_year_type3 <- tibble(
  fit_spotted_year$typeIII.hypoth
) %>% 
  mutate(Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
spotted_year_type3 <- setNames(split(spotted_year_type3, 
                               seq(nrow(spotted_year_type3))), 
                         rownames(spotted_year_type3))
```

The second set of models that included annual variation in haul-out patterns uncovered
significant contributions for linear and quadratic interactions between day and year in
both ribbon seals (day:year,
$F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$ = 
`r ribbon_year_type3$day_year$F.value`; $p$ = `r ribbon_year_type3$day_year$Prob.F`;
day^2^:year,
$F_{`r paste(ribbon_year_type3$day2_year$Num.df, ribbon_year_type3$day2_year$Den.df, sep=",")`}$ = 
`r ribbon_year_type3$day2_year$F.value`; $p$ = `r ribbon_year_type3$day2_year$Prob.F`) and
spotted seals (day:year,
$F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$ = 
`r spotted_year_type3$day_year$F.value`; $p$ = `r spotted_year_type3$day_year$Prob.F`;
day^2^:year,
$F_{`r paste(spotted_year_type3$day2_year$Num.df, spotted_year_type3$day2_year$Den.df, sep=",")`}$ = 
`r spotted_year_type3$day2_year$F.value`; $p$ = `r spotted_year_type3$day2_year$Prob.F`).
Predicted distributions of haul-out activity were largely unimodal, but varied some among
and within years with respect to the timing and magnitude of haul-out peaks (Figure
\@ref(fig:annualHO)). It is important to note that predicted variation in annual haul-out
patterns likely reflects both process error and sampling variability. While we did remove
any years where only one deployment in a species + age-sex group was present, there are
still some years where the pattern shown is informed by a small number of tags that may
not represent population level processes.

@@

The timing of peak haul-out probability for ribbon and spotted seals appeared to have only
a limited to no relationship with the amount of yearly maximum sea ice extent. Adult
female and subadult spotted seals show a negative trend line but no indication that the
observed trend is meaningful (spotted seal adult female: $R^{2}$ =
`r pl_adultF$r.squared`, $p$ = `r pl_adultF$p.value`; spotted seal subadults: $R^{2}$ =
`r pl_subadult$r.squared`, $p$ = `r pl_subadult$p.value`). For ribbon seals and adult male
spotted seals, $p$-values were substantially larger than 0.05 (ribbon seal adult females;
$R^{2}$ = `r hf_adultF$r.squared`, $p$ = `r hf_adultF$p.value`; ribbon seal adult males:
$R^{2}$ = `r hf_adultM$r.squared`, $p$ = `r hf_adultM$p.value`; ribbon seal subadults:
$R^{2}$ = `r hf_subadult$r.squared`, $p$ = `r hf_subadult$p.value`; spotted seals adult
males: $R^{2}$ = `r pl_adultM$r.squared`, $p$ = `r pl_adultM$p.value`).

@@

::: {custom-style="comment"}
This section is unchanged from Paul's original write up. I have put off revisiting and
integrating with my current document/code. Maybe between Paul/Irina/Me we can sort out a
quick update. Might just need to walk me through the previous code and make sure I get it
properly incorporated with this version
:::

@@

In this paper, we modeled data from bio-logging sensors deployed on bearded, spotted, and
ribbon seals to examine factors affecting haul-out behavior on sea ice in the Bering and
Chukchi seas. Our analysis shows seals use sea ice progressively more throughout the
spring, and with peak use near mid-May to early June before declining again. This pattern
aligns well with what has been previously documented and confirms our bio-logging
deployments are likely capturing population level behavioral patterns. Seals
preferentially haul out on ice shortly after solar noon which coincides with the daily
period of maximum solar radiation. The association of daily peak haul-out behavior with
mid to later solar hours is commonly noted in pinnipeds and allows seals to maximize
absorption of solar radiation. Interestingly, bearded seals appear to have two peaks in
haul-out activity across the day, one shortly after solar noon, and one centered near
solar midnight. A similar bi-modal pattern has been seen in ringed seals and suggests that
bearded and ringed seals may be operating under different constraints than ribbon and
spotted seals. Bearded and ringed seals are distributed across higher latitudes and the
extended daylight hours may allow more flexibility in optimizing resting periods with
foraging. Notably, bearded and ringed seals are subjected to much higher levels of
predation from polar bears and human harvest.

::: {custom-style="comment"}
are there references for timing of polar bear activity/kills or hunter activity that might
lend some additional support for predation as a possible explanation for the bi-modal
haul-out pattern? Any data on timing of hunting activity by Alaska communities?
:::

Unlike previous analyses of seal haul-out data (e.g. Ver Hoef et al. 2009, Conn et al.
2014), we also investigated the influence of sex-age class on haul-out probabilities for
all species except bearded seals because of low sample size. While both ribbon and spotted
seals exhibited a unimodal diel haul-out pattern generally centered around local solar
noon, there were key differences across species, age, and sex that match expectations
given what we know of their ecological behavior. Spotted seals are known to form triads
during the breeding season where a female and dependent pup are accompanied on the ice by
a suitor male. The male is waiting for the female to wean the pup and enter estrus and
fending off any other males that may be interested. Triad formation results in both males
and females spending a large portion of the day hauled out on ice and a protracted spring
haul-out season for both sexes. Since breeding occurs in the water, females may also be
less inclined to venture out for foraging trips while still nursing the pup. We can see a
clear reflection of this in the predicted haul-out behavior with both males and females
exhibiting a broad distribution of time out of the water throughout the solar day and the
season. Ribbon seals are not known to form triads and our model predicts a progression of
increased haul-out behavior with females starting earlier in the season than males.
Notably, female ribbon seals spend a large portion of the day in the water during the
pupping period and aligns with the hypothesis that ribbon seal females continue foraging
while nursing. In the case of both ribbon and spotted seals, subadults are the first to
begin meaningful haul-out behavior and reflects a typical phocid pattern where subadults
molt first as they do not have any reproductive constraints.

::: {custom-style="comment"}
this previous paragraph might benefit from some references e.g. triads in spotted seals,
ribbon seal females foraging while nursing, etc paper on dive/foraging behavior of
lactating harp seals: <https://doi.org/10.1006/anbe.1993.1312>
:::

We also investigated the influence of environmental covariates on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind chill.
These have been investigated for walrus (e.g. Udevitz et al. [@udevitz2009cjz]), but to
our knowledge have never been investigated for ice-associated seals. Ribbon seals seem to
be the most influenced by weather with wind, temperature, barometric pressure, and
precipitation all being important components of the model. Spotted seals were most
affected by wind and barometric pressures. And, for bearded seals, the model indicated
wind and temperature had the greatest impact. In general, and as might be expected, seals
were more likely to haul-out when daily temperatures were warmer, winds speeds were less,
barometric pressure was high, and lower rates of precipitation. These weather conditions
are likely most beneficial for energetic considerations and may also provide the best
situation for predator detection. These results highlight the importance for haul-out
behavior models (and subsequent calculations of availability during surveys) to
incorporate weather covariates.

Our model uncovered annual deviations in the timing and magnitude of
haul-out peaks for ribbon and spotted seals. The timing of peak haul-out activity appears
to fall within a relatively narrow time window of 3-4 weeks in late May and early June. 
This remarkable level of consistency across 15 years is indicative of the 
adaptations these species have evolved to synchronize their key life history stages
with favorable environmental conditions (e.g. availability of sea ice, increased
oceanographic productivity, favorable weather conditions). Unfortunately, this may also
be indicative of a limited range of plasticity as climatological conditions change.
Additionally, while from an ecological perspective the haul-out behavior appears
consistent, the interannual differences in timing and magnitude are large enough to have
important ramifications on calculations of abundance and trend.

Previous attempts to estimate the abundance of phocid seals from aerial survey data in the
Bering and Chukchi Seas (e.g. Bengtson et al. 2005, Conn et al. 2014, Ver Hoef et al.
2014) have used estimated haul-out probabilities to correct for the proportion of animals
that are in the water and thus unavailable to be counted during aerial surveys. Although
several of these studies allowed haul-out probabilities to vary by day-of-year and
time-of-day, previous correction factors have not accounted for variability between years,
different weather conditions, and the age-sex class of the sample. In this paper, we have
shown that there can be considerable differences in the number of seals hauled out on ice
based on these factors. We thus recommend that future abundance analyses employ correction
factors that account for these factors. For instance, it is relatively straightforward to
obtain weather reanalysis products for times and locations that are surveyed and to
construct a relevant correction factor based on predictions of GMPMs. Further, one can
apply stable stage distribution estimates from matrix population models to adjust
predictions to the "average" prediction of the population. The latter suggestion is
particularly appropriate when the age structure of the tagged sample does not reflect the
likely age structure of the population. For instance, population weighted estimates can
differ substantially from those based on a naïve analysis of the tagged sample, as was the
case with analysis of spotted seal data (e.g. Fig X). The most challenging element in
developing availability correction factors is with annual variability. It can be difficult
to get a sufficient sample size to estimate year-specific correction factors, particularly
because research teams would likely need to tag seals and conduct aerial surveys
concurrently which requires considerably more personnel and money. One possible suggestion
is to estimate a "shift" parameter within models for aerial survey counts that allow the
peak of haul-out distributions to be adjusted earlier or later in the year based on the
frequency of counts observed over time. Regardless, researchers should anticipate there
being some un-modeled heterogeneity in availability probability present in abundance
estimates obtained from aerial surveys. This may make trend detection difficult, as one
will not know if moderate differences in abundance estimates are attributable to changes
in abundance or changes in haul-out frequencies.

Predictions of absolute haul-out probability in this paper were somewhat different than
those previously reported for these species, especially for bearded seals. For instance,
Ver Hoef et al. (2014) and Conn et al. (2015) used haul-out correction factors with
maximums of 0.66 for bearded seals, 0.62 for ribbon seals, and 0.54 for spotted seals,
where maximums corresponded to times near solar noon in mid-late May. Applying models that
ignore age, sex, and year effects, these probabilities were 0.38, 0.72, and 0.60,
respectively, under the current analysis framework. Our current estimates reflect
increased sample sizes in terms of number of animals, but also improvements to the way
data are prepared prior to analysis.
